---
title: "Midterm_1_cheatsheet"
output: html_document
date: "2023-01-31"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

Run these!!

```{r cars}
library(janitor)
library(tidyverse)
library(skimr)
```

## Lab 2.1 Functions

#Types of Data Use the 'class()' function to see what type of data you
are working with

```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
my_complex <- 2+4i
```

The `is()` and `as()` functions to clarify or specify a type of data.

```{r}
is.integer(my_numeric) #is my_numeric an integer?

my_integer <- 
  as.integer(my_numeric) #create a new object specified as an integer

is.integer(my_integer) #is my_numeric an integer?
```

Example: change type of data to a factor using as.factor()

```{r}
fisheries$country <- as.factor(fisheries$country)
```

Missing Data (NAs) Useful functions when dealing with NAs: `is.na` or
`anyNA`

```{r}
my_missing <- NA

is.na(my_missing) #is na my missing value 

anyNA(my_missing)
```

Example:

```{r}
new_vector <- c(7, 6.2, 5, 9, NA, 4, 9.8, 7, 3, 2)

mean(new_vector)

mean(new_vector, na.rm=T) #na.rm removes the NA values in the vector
#T means true 
```

##Lab 2.2 Functions

#Vectors - Organize data in R - Create vectors using the `c` command -
The `c` stands for concatenate.

```{r}
#numeric vector 
my_vector <- c(10, 20, 30)

#character vector (always has quotes)
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thrusday", "Friday", "Saturday", "Sunday")

#trick for generating a sequence of numbers 
my_vector_sequence <- c(1:100)
my_vector_sequence
```

# Identifying vector elements

-   use `[]` to pull out elements in a vector
-   specifiy their position in the vector

```{r}
days_of_the_week[3] #to get Wednesday 
```

# Data Matrices

-   Data matrices are a series of stacked vectors, similar to a data
    table

```{r}
Philosophers_Stone <- c(317.5, 657.1)
Chamber_of_Secrets <- c(261.9, 616.9)
Prisoner_of_Azkaban <- c(249.5, 547.1)
Goblet_of_Fire <- c(290.0, 606.8)
Order_of_the_Phoenix <- c(292.0, 647.8)
Half_Blood_Prince <- c(301.9, 632.4)
Deathly_Hallows_1 <- c(295.9, 664.3)
Deathly_Hallows_2 <- c(381.0, 960.5)
```

Create new object to combine all the vectors into one (using 'c'
command)

```{r}
box_office <- c(Philosophers_Stone, Chamber_of_Secrets, Prisoner_of_Azkaban, Goblet_of_Fire, Order_of_the_Phoenix, Half_Blood_Prince, Deathly_Hallows_1, Deathly_Hallows_2)
box_office
```

Create matrix using `matrix()` function - use the `nrow` and `byrow`
commands to organize the matrix

```{r}
harry_potter_matrix <- matrix(box_office, nrow = 8, byrow = T)
harry_potter_matrix
```

#Name the rows and columns

1.  Create vectors for the titles of the rows and columns

```{r}
region <- c("US", "non-US")
region

titles <- c("Philosophers_Stone", "Chamber_of_Secrets", "Prisoner_of_Azkaban", "Goblet_of_Fire", "Order_of_the_Phoenix", "Half_Blood_Prince", "Deathly_Hallows_1", "Deathly_Hallows_2")
titles
```

2.  Columns `colnames()`

```{r}
colnames(harry_potter_matrix) <- region
```

3.  Rows `rownames()`

```{r}
rownames(harry_potter_matrix) <- titles
```

4.  Print Matrix

```{r}
harry_potter_matrix
```

#Analysis of a Data Matrix

Calculate 'rowSums'

```{r}
global <- rowSums(harry_potter_matrix)
global
```

Add columns `cbind()`

```{r}
all_harry_potter_matrix <- cbind(harry_potter_matrix, global)
all_harry_potter_matrix
```

Selecting Elements from a Data Matrix - Use `[,]` [Row,Column]

```{r}
harry_potter_matrix[2,1] #selects the value in the first column, second row
```

-   Adding a colon `:` selects the specified elements in a column.

```{r}
harry_potter_matrix[1:4]
```

## Lab 3.1 Functions

#Data Frames - The most common way to organize data within R - Similar
to a spreadsheet - Can store multiple classes of data

1.  Make seperate vectors

```{r}
Sex <- c("male", "female", "male")
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)
```

2.  Combine the vectors into a data frame using `data.frame()`

```{r}
hbirds <- data.frame(Sex, Length, Weight)
hbirds
```

#Analysis of Data Frames

-   Names of columns: 'names(data frame name)'
-   Dimensions: `dim()` and `str()`
-   Renaming to lowercase

```{r}
hbirds <- data.frame(sex = Sex, length_in = Length, weight_oz = Weight) #renaming will become more helpful in later labs
names(hbirds)
```

-   Values from an entire column using the `$` sign

```{r}
w <- hbirds$weight_oz
mean(w)
```

-   Adding rows using the `rbind()` function

```{r}
new_bird <- c("female", 3.6, 3.9)
new_bird

hbirds<- rbind(hbirds, new_bird)
hbird
```

-   Adding columns/ more data in the data frame

```{r}
# Use the `$` operator
hbirds$neighborhood <- c("lakewood", "brentwood", "lakewood", "scenic Heights")
hbirds
```

#CSV Files - Use `row.names = FALSE` to avoid row numbers from printing
out

```{r}
write.csv(hbirds, "hbirds_data.csv", row.names = FALSE)
```

## How to load in new data

```{r}
library("tidyverse")
```

```{r}
bison <-readr::read_csv("data/bison.csv")
```

## Lab 3.2 Functions

-   Show levels of the factor

```{r}
levels(hot_springs$scientist)
```

# Summary functions

We can summarize our data frame with the`summary()` function.

```{r}
summary(fish)
```

`glimpse()` is another useful summary function.

```{r}
glimpse(fish)
```

`nrow()` gives the numbers of rows.

```{r}
nrow(fish) #the number of rows or observations
```

`ncol` gives the number of columns.

```{r}
ncol(fish) #the number of columns or variables
```

`dim()` gives the dimensions.

```{r}
dim(fish) #total dimensions
```

`names` gives the column names.

```{r}
names(fish) #column names
```

`head()` prints the first n rows of the data frame.

```{r}
head(fish, n = 10)
```

`tail()` prints the last n rows of the data frame.

```{r}
tail(fish, n = 10)
```

`table()` is useful when you have a limited number of categorical
variables. It produces fast counts of the number of observations in a
variable. We will come back to this later...

```{r}
table(fish$lakeid)
```

We can also click on the `fish` data frame in the Environment tab or
type View(fish).

```{r}
View(fish)
```

# Filter

Filter is a way of pulling out observations that meet specific criteria
in a variable. We will work a lot more with this in the next lab.

```{r}
little_fish <- filter(fish, length<=100)
little_fish
```

## Lab 4.1 Functions

Tidyverse - an "opinionated" collection of packages that make workflow
in R easier. - observations (rows) - filter - variables (columns) -
select - values

Select (`select()`) - allows you to pull out columns of interest from a
dataframe

```{r}
select(fish, "lakeid", "scalelength")
```

To add a range of columns use `start_col:end_col`

```{r}
select(fish, fish_id:length)
```

The - operator is useful in select. It allows us to select everything
except the specified variables

```{r}
select(fish, -fish_id, -annnumber, -length, -radii_length_mm)
```

-   Varaibles that contain a certain value of interest

```{r}
select(fish, contains("length"))
```

Options to select columns based on a specific criteria include:\
1. ends_with() = Select columns that end with a character string\
2. contains() = Select columns that contain a character string\
3. matches() = Select columns that match a regular expression\
4. one_of() = Select columns names that are from a group of names\
5. starts_with()

To select all columns that are *not* a class of data, you need to add a
`~`.

```{r}
select_if(fish, ~!is.numeric(.))
```

When naming columns, blank spaces are often added. Here is a trick to
remove these.

```{r}
select_all(mammals, ~str_replace(., " ", "_"))
```

## Lab 4.2 Functions

#Filter

-   `filter()` allows all of the expected operators; i.e. \>, \>=, \<,
    \<=, != (not equal), and == (equal).\

```{r}
filter(fish, lakeid == "AL")

filter(fish, length >= 350)

filter(fish, lakeid != "AL")
```

-   Filtering multiple values within the same variable requires the
    `%in%`

```{r}
filter(fish, length %in% c(167, 175))
```

-   Use `between` if you are looking for a range of specific values

```{r}
filter(fish, between(scalelength, 2.5, 2.55))
```

-   Extract observations "near" a certain value but you need to specify
    a tolerance

```{r}
filter(fish, near(radii_length_mm, 2, tol = 0.2))
```

The '&' operator - Extract only the fish that have lakeid "AL" and
length \>350.

```{r}
filter(fish, lakeid == "AL" & length > 350)
```

The `|` operator - Extracts either of the conditons

```{r}
filter(fish, lakeid == "AL" | length > 350)
```

Rules:\
+ `filter(condition1, condition2)` will return rows where both
conditions are met.\
+ `filter(condition1, !condition2)` will return all rows where condition
one is true but condition 2 is not.\
+ `filter(condition1 | condition2)` will return rows where condition 1
or condition 2 is met.\
+ `filter(xor(condition1, condition2)` will return all rows where only
one of the conditions is met, and not when both conditions are met.

##Lab 5.2 Functions

#Mutate function (shift+command+m)

`mutate()`

```{r}
#Changing many variables to factors
mammals %>% mutate_if(is.character, as.factor)
```

-   Allows us to create a new column from existing columns in a data
    frame.

```{r}
fish %>% 
  mutate(length_mm = length*10) %>% 
  select(fish_id, length, length_mm)
```

`mutate_all()` - Helpful when cleaning data - Mixed entries (upper and
lowercase), blank spaces, odd characters, etc

```{r}
mammals %>%
  mutate_all(tolower)
```

-   Specify individual columns

```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower))
```

#Pipes `%>%` Combines `select()`, `filter()`, and other functions - Feed
the output from one function into the input of another function

```{r}
fish %>%
  select(lakeid, scalelength) %>%
  filter(lakeid == "AL")
```

Or to update the object:

```{r}
#fish <- 
#fish %>% 
  #select(lakeid, scalelength) %>%
  #filter(lakeid == "AL")
```

#Arrange `arrange()` - Default is ascending order

```{r}
fish %>% 
  select(lakeid, scalelength) %>% 
  arrange(scalelength)
```

-   Sort in decreasing order, wrap the variable name in `desc()`

```{r}
fish %>% 
  select(lakeid, scalelength) %>% 
  arrange(desc(scalelength))
```

# If Else `if_else()`

-   First specify a logical statement, afterwards what needs to happen
    if the statement returns `TRUE`, and lastly what needs to happen if
    it's `FALSE`.

```{r}
#use `if_else()` to replace `-999.00` with `NA`.
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  arrange(newborn)
```

##Lab 5.2 Functions

#Janitor - `clean_names` function

```{r}
#library("janitor")

superhero_powers <- janitor::clean_names(superhero_powers)
```

# `tabyl`

-Produces counts and percentages

##Lab 6.1 Functions

#Skimr package

```{r}
#install.packages("skimr")
library("skimr")
```

Try out the new function `skim()` as part of the `skimr` package.
##gives you multiple outputs of the data

```{r}
skim(msleep24)
```

Histograms are also a quick way to check the output.

```{r}
hist(msleep24$sleep_total_24)
```

#Summarize `summarize()`

-We can accomplish the same task using the `summarize()` function to
make things cleaner.

```{r}
msleep %>% 
  filter(bodywt > 200) %>% #large mammals
  summarize(mean_sleep_lg = mean(sleep_total))
```

You can also combine functions to make useful summaries for multiple
variables.

```{r}
msleep %>% 
    filter(bodywt > 200) %>% 
    summarize(mean_sleep_lg = mean(sleep_total), 
              min_sleep_lg = min(sleep_total),
              max_sleep_lg = max(sleep_total),
              total = n())
```

```{r}
names(msleep)
```

# `n_distinct()`

-   A very handy way of cleanly presenting the number of distinct
    observations

```{r}
#Here we show the number of distinct genera over 100 in body weight.

msleep %>% 
  filter(bodywt > 100) %>% 
  summarise(n_genera=n_distinct(genus))
```

#Useful Summary Statistics - sd(), min(), max(), median(), sum(), n()
(returns the length of a column), first() (returns first value in a
column), last() (returns last value in a column) and n_distinct()
(number of distinct values in a column).

## `group_by()`

The `summarize()` function is most useful when used in conjunction with
`group_by()`

```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology, a categorical variable
  summarize(min_bodywt = min(bodywt),
            max_bodywt = max(bodywt),
            mean_bodywt = mean(bodywt),
            total=n())
```

##Lab 6.2 Functions

# Counts

-   `count()` is an easy way of determining how many observations you
    have within a column. It acts like a combination of `group_by()` and
    `n()`

```{r}
penguins %>% 
  count(island, sort = T) #sort=T sorts the column in descending order
```

You can also use `count()` across multiple variables.

```{r}
penguins %>% 
  count(island, species, sort = T) # sort=T will arrange in descending order
```

# Across `across()`

-   There is a function in dplyr called `across()` which is designed to
    work across multiple variables

```{r}
#By using `across()` we can reduce the clutter and make things cleaner
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```
