---
title: "Lab 9 Intro"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
```

## Seating
1. If you already have a group, it will be helpful to start sitting next to each other.
2. If you don't have a group, hang tight we are working on it.
3. Please setup your computer as normal for today.

## Warm-up
1. In the data folder there is an epidemiology data set on an outbreak of malaria.
2. `rdt` refers to rapid diagnostic test and they are identified here by age group.
3. Make the data tidy and store them as a new object.
4. Which district had the highest *total* number of cases on July 30, 2020?

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor)
```


Malaria csv file 
```{r}
malaria <- read_csv("data/malaria.csv") %>% clean_names()
malaria
```

Pivot longer 
```{r}
malaria_long <- malaria %>% 
  pivot_longer(cols=starts_with("malaria_rdt"),
               names_to = "age_class",
               values_to = "cases") %>% 
  select(newid, data_date, submitted_date, location_name, province, district, age_class, cases)

malaria_long
```
```{r}
malaria_long %>% 
  filter(data_date == "2020-07-30") %>% 
  group_by(district) %>% 
  summarize(tot_cases= sum(cases, na.rm = T)) %>% 
  arrange(-tot_cases)
```

