---
title: "Midterm_2_cheatsheet"
output: html_document
date: "2023-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

here ya go!

```{r}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
library(RColorBrewer)
library(paletteer)
library(skimr)
library(palmerpenguins)
library(gtools)
```

# Lab 7 - NA's

## Replacing -999

Use `na_if`

```{r}
amniota_tidy <- amniota %>% 
  na_if("-999")
```

Treating recorded data as NA's

```{r}
msleep %>%  
  mutate(conservation_modified = na_if(conservation, "domesticated"))%>% 
  count(conservation_modified, sort = T)
```

## naniar

`miss_var_summary` provides a nice, clean summary of NA's across the data frame.

```{r}
naniar::miss_var_summary(amniota_tidy)
```

------------------------------------------------------------------------

# Lab 8 - Pivot_longer, Pivot_wider, Separate

```{r}
heartrate <- read_csv("data/heartrate.csv")
```

## Pivot longer

Makes column names into data

`pivot_longer()` makes a new column using the column names and brings data from frame into a new column

```{r}
heartrate #before pivoting
heartrate %>% 
  pivot_longer(-patient, #will not pivot
               
               names_to = "drug", #Takes column names and makes it a new column                                   with the name you choose
               
               values_to = "heartrate" #values from the columns you pivoted                                             become a new column with name you choose
               )
```

To choose the range of columns that you want to pivot, do one of the following:

Solution 1: specify a range of columns that you want to pivot.

```{r}
billboard <- readr::read_csv("data/billboard.csv")
billboard2 <- 
  billboard %>% 
  pivot_longer(wk1:wk76, # a range of columns
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE #this will drop the NA's
               )
#billboard2
```

Solution 2: OR, specify columns that you want to stay fixed.

```{r}
billboard3 <- 
  billboard %>% 
  pivot_longer(-c(artist, track, date.entered), #specific columns
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE
               )
#billboard3
```

Solution 3: identify columns by a prefix, remove the prefix and all NA's.

```{r}
billboard4 <- billboard %>% 
   pivot_longer(
   cols = starts_with("wk"),
   names_to = "week",
   names_prefix = "wk",
   values_to = "rank",
   values_drop_na = TRUE)
#billboard4
```

## Separate and unite

Use `separate()` to separate columns via their names

`separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.

```{r}
heartrate2 <- readr::read_csv("data/heartrate2.csv")
heartrate2 #we want to separate patient names from their sexes

heartrate3 <- heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_")
heartrate3
```

`unite()` is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.

```{r}
heartrate3 %>% 
  unite(patient_sex, "patient", "sex", sep = " ")
```

## Pivot wider

You use `pivot_wider()` when you have an observation scattered across multiple rows.

Rules:\
+ `pivot_wider`(names_from = "", values_from ="")\
+ `names_from` - Values in the `names_from` column will become new column names + `values_from` - Cell values will be taken from the `values_from` column

```{r}
tb_data <- read_csv("data/tb_data.csv")
tb_data #split the key column values into their own columns

tb_data %>% 
  pivot_wider(names_from = "key", #the observations under key will become new columns
              values_from = "value")
```

What happens when we want to use `pivot_wider()` with multiple columns?

```{r}
edu_level <- readr::read_csv("data/education_level.csv")
edu_level
```

We want to pivot the `edu_level` data such that there is only a single row per demographic.

```{r}
edu_level %>% 
pivot_wider(names_from = (education_level), #new column names come from the education_level column
                values_from = c(mean_income, n)) #values come from two separate columns
```

# Lab 9 - ggplot part 1

## Data Types

We first need to define some of the data types we will use to build plots.

-   `discrete` quantitative data that only contains integers EX: numbers of things
-   `continuous` quantitative data that can take any numerical value EX: weights, heights
-   `categorical` qualitative data that can take on a limited number of values EX: sex

## ggplot basics

The basic syntax is in short, **plot= data + geom\_ + aesthetics**.

```{r}
ggplot(data=iris, mapping=aes(x=Species, y=Petal.Length))+geom_boxplot()
```

## Scatterplots and barplots

Scatter plots allow for comparisons of two continuous variables.

Scatterplots use `geom_point()`

```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
ggplot(homerange, mapping=aes(x=log10.mass, y=log10.hra))+geom_point()+
  geom_smooth(method=lm, se=T) #add a line of best fit
```

Barplots use either `geom_bar()` or `geom_col()`

```{r}
homerange %>% 
  ggplot(aes(x=trophic.guild)) + 
  geom_bar() #good for counts
```

`geom_col()` allows us to specify an x-axis and a y-axis.

```{r}
homerange %>% 
  filter(family=="salmonidae") %>%
  select(common.name, log10.mass) %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_col()
```

## Boxplots

Boxplots help us visualize a range of values. So, on the x-axis we typically have something categorical and the y-axis is the range.

Boxplots use `geom_boxplot`

```{r}
homerange %>% 
  ggplot(aes(x=class, y = log10.mass))+
  geom_boxplot()
```

# Lab 10 - ggplot part 2

## coord flip

Use `coord_flip()` to make the barplot horizontal (and easier to read)

```{r}
life_history <- read_csv("data/mammal_lifehistories_v2.csv", na="-999") %>% clean_names()

life_history %>% 
  count(order, sort=T) %>% 
  ggplot(aes(x=order, y=n))+ geom_col()+ coord_flip()
```

## scale_y\_log10

Use `scale_y_log10()` when the y values are too spread out

```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T)) %>% 
  ggplot(aes(x=order, y=mean_mass))+ geom_col()+
  coord_flip()+
  scale_y_log10()
```

We can also scale all bars to a percentage (or proportion).

```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild))+
  geom_bar(position = position_fill())+ 
  scale_y_continuous(labels = scales::percent)+
  coord_flip()
```

## Aesthetics: Labels

To add custom labels to the axis or title, we use the `labs` command.

```{r}
elephants <- read_csv("data/elephantsMF.csv") %>% clean_names()

elephants %>% 
  ggplot(aes(x=age, y=height)) + 
  geom_point() + 
  geom_smooth(method=lm, se=F)+
  labs(title = "Elephant Age vs Height", 
       x="Age",
       y="Height")
```

We can improve the plot further by adjusting the size and face of the text. We do this using `theme()`.

```{r}
elephants %>% 
  ggplot(aes(x=age, y=height)) + 
  geom_point() + 
  geom_smooth(method=lm, se=F)+
  labs(title = "Elephant Age vs Height", 
       x="Age",
       y="Height")+
  theme(plot.title = element_text(size=rel(1.5), hjust=0.5))
```

The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` allows control of title position.

```{r}
elephants %>% 
  ggplot(aes(x=age, y=height)) + geom_point() + geom_smooth(method=lm, se=F) +
  labs(title = "Elephant Age vs. Height",
       x = "Age",
       y = "Height") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```

We can tilt the labels to make the graph easier to read

```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) 
```

## Aesthetics: Point sizes and Shapes

Put `size =` inside the geom to change the size of the point

```{r}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x= log10.mass, y= log10.hra))+
  geom_point(size=1)
```

Map shapes to another categorical variable using `aes()` inside the geom

```{r}
homerange %>% 
  ggplot(aes(x= log10.mass, y= log10.hra))+
  geom_point(aes(shape=thermoregulation, color=thermoregulation), size=1.75)
```

## Fill

Use `fill` to group (what you say `fill =` should be a categorical data)

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) + geom_bar() +
  coord_flip() +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill = "Trophic Guild")
```

We can also have counts of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) + geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill = "Trophic Guild")
```

## Group

In addition to `fill`, `group` is an aesthetic that accomplishes the same function but does not add color.

Here is a box plot that shows `log10.mass` by taxonomic `class.`

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = class, y = log10.mass)) +
  geom_boxplot()
```

I use `group` to make individual box plots for each taxon.

```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass, group = taxon)) +
  geom_boxplot()
```

I can also use `fill` to associate the different taxa with a color coded key.

```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10.mass, fill = taxon)) +
  geom_boxplot()
```

# Lab 11 - ggplot part 3

## Line plots

Line plots are great when you need to show changes over time.

Line points are created by using `geom_line()`

Here we look at the number of samples for species DM and DS over the years represented in the data.

```{r}
deserts <- read_csv("data/surveys_complete.csv")

deserts %>% 
  filter(species_id=="DM" | species_id=="DS") %>% 
  group_by(year, species_id) %>% 
  summarise(n=n(), .groups='keep')
```

```{r}
deserts2 <- deserts %>% mutate(year=as_factor(year))
```

```{r}
deserts2 %>% 
  filter(species_id=="DM" | species_id=="DS") %>% 
  group_by(year, species_id) %>% 
  summarise(n=n(), .groups='keep') %>% 
  ggplot(aes(x=year, y=n, group=species_id, color=species_id))+
  geom_line()+
  geom_point(shape=5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Number of samples for species DM & DS",
       x = "Year",
       fill = "n")
```

## Histograms

Histograms show the distribution of continuous variables.

A histogram `bins` the data and you specify the number of bins that encompass a range of observations. By default, R uses a formula to calculate the number of bins but some adjustment is often required.

What does the distribution of body mass look like in the `homerange` data?

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(alpha = 0.3, color = "black", fill = "purple", bins=200)+ 
  labs(title = "Distribution of Body Mass")
```

We can us `alpha` inside the geom to change the transparency of the graph

To get a list of colors:

```{r}
grDevices::colors()
```

## Density plots

Density plots are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_density(fill="deepskyblue4", alpha  =0.4, color = "black")+
  labs(title = "Distribution of Body Mass")
```

I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.

```{r message=FALSE}
homerange %>% 
  ggplot(aes(x=log10.mass)) +
  geom_histogram(aes(y = ..density..), fill = "deepskyblue4", alpha = 0.4, color = "black")+
  geom_density(color = "red")+
  labs(title = "Distribution of Body Mass")
```

## Create Categories with mutate and case_when()

`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions.

We can use case_when() to make a new column `range_category` that breaks down `log10.hra` into very small, small, medium, and large classes based on quartile.

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

summary(homerange$log10.hra)
```

```{r}
library(gtools)
quartiles <- quantcut(homerange$log10.hra)
table(quartiles)
```

```{r}
homerange <- homerange %>% 
  mutate(range_category = case_when(log10.hra<=3.653 ~ "very_small",
                                    log10.hra >=3.653 & log10.hra <= 4.595 ~ "small",
                                    log10.hra >= 4.595 & log10.hra <=6.016 ~ "medium",
                                    log10.hra >= 6.016 ~"large"))
homerange
```

## ggplot themes

There are many options to change the theme of your plots. Have a look [here](https://www.datanovia.com/en/blog/ggplot-themes-gallery/) for some excellent examples.

Here is a list of the `ggthemes`
```{r}
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]
```

```{r}
#homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")

homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge", alpha = 0.85)+
  theme_light()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Legends

There are lots of options to manipulate legends. Have a look [here](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software).
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge", alpha = 0.85)+
  theme_linedraw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```


## RColorBrewer

Makes your stuff look nicer!


The thing to notice is that there are three different palettes: 1) sequential, 2) diverging, and 3) qualitative. Within each of these there are several selections. You can bring up the colors by using `display.brewer.pal()`. Specify the number of colors that you want and the palette name.
```{r}
display.brewer.pal(4,"GnBu")
```

The [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) website is very helpful for getting an idea of the color palettes. To make things easy, use these two guidelines:

+`scale_colour_brewer()` is for points  
+`scale_fill_brewer()` is for fills  

Here's an example of how to use it:
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge", alpha = 0.85)+
  scale_fill_brewer(palette = "PuRd")+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Manually Setting Colors
You can also use `paleteer` to build a custom palette for consistency. To access the `paleteer` collection, I add it to a new object.
```{r}
colors <- paletteer::palettes_d_names
colors
```

Now we can display the palettes. Assign the palette to `my_palette` and then build this base R bar plot. There are a lot of options; `paleteer` is a collection of popular palettes. I really like the [`ggsci` package] (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
```{r}
my_palette <- paletteer_d("colorBlindness::paletteMartin")
```

```{r}
barplot(rep(1,14), axes=FALSE, col=my_palette)
```

Here's an example of how to use it:
It changes to `scale_fill_manual()`
```{r}
homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge", alpha = 0.85)+
  scale_fill_manual(values=my_palette)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```


## Adjusting the x and y limits

To adjust limits, we can use the `xlim` and `ylim` commands. When you do this, any data outside the specified ranges are not plotted.

```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point() +
  xlim(0, 4) +
  ylim(1, 6)
```


## Faceting

Faceting allows us to make multi-panel plots for easy comparison.  

Import the data
```{r}
ecosphere <- read_csv("data/ecs21351-sup-0003-SupplementS1.csv", skip=2, col_names = T) %>% 
  clean_names() %>% 
  separate(scientific_name, into = c("genus", "species"), sep=" ") %>% 
  na.omit() # remove NA's for now
```

`facet_wrap()` makes a ribbon of panels, but you can control how you want them arranged.  
```{r}
ecosphere %>% 
  ggplot(aes(x=diet, y=log10_mass, fill=migratory_strategy))+ 
  geom_boxplot(alpha=0.4) + 
  facet_wrap(~migratory_strategy, ncol=4)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Mass by Diet and Migratory Strategy",
       x = NULL,
       y = "Log 10 Mass",
       fill = "Migratory Strategy")
```

`facet_grid()` allows control over the faceted variable; it can be arranged in rows or columns.  
rows~columns
```{r}
ecosphere %>% 
  ggplot(aes(x=diet, y=log10_mass, fill=migratory_strategy))+ 
  geom_boxplot(alpha=0.4)+ 
  facet_grid(migratory_strategy~.)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Mass by Diet and Migratory Strategy",
       x = NULL,
       y = "Log 10 Mass",
       fill = "Migratory Strategy")
```

`facet_grid()` will also allow the comparison of two categorical variables, just remember a~b where a is rows and b is columns.  
```{r}
ecosphere %>% 
  ggplot(aes(x=log10_mass))+
  geom_density()+
  facet_grid(diet~habitat, scales = "free_y")
```


